---
title: "HW 03"
author: "Lucas Smith"
format:
  html:
    embed-resources: true
toc: true
---

```{r}
pkgs = c("tidyverse", "grid", "gridExtra", "ggplot2", "reshape", "ggalt", "likert", "ggstats")
for (pkg in pkgs) {
  if (!require(pkg, character.only=TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```

## 1 - Du Bois challenge.

```{r}
#| label: label-me-1

dubois <- read_csv("data/income.csv")
db <- pivot_longer(dubois, 
                   c("Rent", "Food", "Clothes", "Tax", "Other"), 
                   names_to="income_type", values_to = "percent")
order <- rev(c("$100-200", 
               "$200-300", 
               "$300-400", 
               "$400-500", 
               "$500-750", 
               "$750-1000", 
               "$1000 AND OVER"
               )
             )
color_vals = c("Rent" = "#141512", "Food" = "#805D7D", 
               "Clothes" = "#DA907F", "Tax"="#958c82", 
               "Other"="#BDB39C"
               )
x_ordering <- c("Rent", "Food", "Clothes", "Tax", "Other")

plt <- ggplot(db, 
              aes(x=percent, y=Class, fill=factor(income_type, levels=rev(x_ordering)))) +    geom_col(position="fill") 
plt <- plt + 
        guides(fill=FALSE) + 
        theme(axis.text.x=element_blank(),
             axis.title.x=element_blank(),
             axis.ticks.x=element_blank(), 
             axis.ticks.y = element_blank(),
             legend.position = NULL, 
             plot.margin=margin(0,0,0,0),
             plot.background = element_blank(),
             axis.text.y = element_blank(),
             panel.background=element_blank()) + 
        labs(x="", y="") +
        scale_y_discrete(limits=order) +
        scale_fill_manual(values=color_vals) 

tabledata <- dubois |> select(Class, Average_Income)
tab = tableGrob(tabledata, rows=NULL, theme=ttheme_default(base_size = 7, ))
tab$heights <- unit(rep(1/nrow(tab), nrow(tab) ), "npc")

grid.arrange(
  arrangeGrob(tab, nullGrob(), heights=c(200, 5)),
  arrangeGrob(nullGrob(), plt, heights=c(4, 30)),
  widths=c(3, 8)
)
```

## 2 - COVID survey - interpret

```{r}
#| label: label-me-2

```

First I would like to mention that I do not like the Likert scaling used in the questionnaire. Most Likert scales have 5 as `Strongly Agree`, yes for this one it is 1. This provides a difference in the user interaction and user experience, and can cause some confusion on reporting as it is an inconsistent experience when compared to other standardized questions that use the Likert scale. Anyway.

There is a lot going on with this visualization. When I saw this visualization for the first time, I felt overwhelmed at the amount of information, and I believe it is important to understand why.

-   The first thing that I notice is the challenge in deciphering any given point. There are three dimensions shown in the graph: the question, the stratification of individual, and the sub-stratification. To view one range, you need to carefully look back and forth a few times to ensure you are on the right scale, which presents a high cognitive load.

-   The second question I noticed is the challenge to discern anything useful. There is so much data, yet at the same time the question becomes: how is any of this useful?

As for the results:

-   Most scores have a low range, from 1-2.

-   To answer the question of whether I believe it is in line with my intuition, I say: it depends. As a general belief system, I believe in the scientific community and as seeing the good in others as well. I believe that most people thought the vaccines were good and were a benefit to society. That part of me does want to go back to the Likert scaling inversion I mentioned before, and say that most people did not read the specifics of what *this* questionnaire's 1 and 5 meant, and assumed that 1 meant disagree and 5 meant agree.

-   The other part of me thinks that it is feasible that many people were skeptical of the vaccine at the time it was released. It was released under different circumstances than a regular vaccine, and I do remember skepticism at the time.

Examples

-   All people thought that they did trust information about the vaccines. The lowest score was a 2.

-   Similarly, all nursing professionals stated that they would recommend the vaccine to friends/family as a 1; in this scoring I would assume that to be accurate.

-   The final note, all medical professionals had a mean value of less than 2 when asked if the vaccine was safe. I would also assume this to be accurate in this scoring.

## 3 - COVID survey - reconstruct

```{r}
covid <- read_csv("data/covid-survey.csv")
```

```{r}
covid <- read_csv("data/covid-survey.csv", skip=1)

```

```{r}
# filter out any no longer useful
colnames <- colnames(covid[, -1])
filtered_missing_rows <- covid |> filter(!if_all(.cols=colnames, is.na))
dim(filtered_missing_rows)
```

-   Relabel the survey response values according to the information in `data/covid_survey.pdf`. Peeking at the plot you’re working towards reconstructing can also be helpful to identify how exactly to recode the variables. Print the dimensions of the resulting data frame with `dim()`.

    -   `exp_already_vax` and `exp_flu_vax`: 0 - No, 1 - Yes

    -   `exp_profession`: 0 - Medical, 1 - Nursing

    -   `exp_gender`: 0 - Male, 1 - Female, 3 - Non-binary third gender, 4 - Prefer not to say

    -   `exp_race`: 1 - American Indian / Alaskan Native, 2 - Asian, 3 - Black / African American, 4 - Native Hawaiian / Other Pacific Islander, 5 - White

    -   `exp_ethnicity`: 1 - Hispanic / Latino, 2 - Non-Hispanic/Non-Latino

    -   `exp_age_bin`: 0 - \<20, 20 - 21-25, 25 - 26-30, 30 - \>30

```{r}
survey_response_values <- c(
  "1" = "Strongly agree",
  "2" = "Somewhat agree",
  "3" = "Neither agree nor disagree",
  "4" = "Somewhat disagree",
  "5" = "Strongly disagree"
)

relabeled_cols <- colnames(covid[, seq(from=9, to=14)])
covid2 <- covid

covid <- covid |> mutate(across(relabeled_cols, ~ survey_response_values[as.character(.)]))
covid <- covid |>
  mutate(exp_already_vax=case_when(
    exp_already_vax== 0 ~ "No", exp_already_vax== 1 ~ "Yes"
  )) |>
  mutate(exp_flu_vax=case_when(
    exp_flu_vax== 0 ~ "No", exp_flu_vax== 1 ~ "Yes"
  )) |>
  mutate(exp_profession=case_when(
    exp_profession== 0 ~ "Medical", exp_profession== 1 ~ "Nursing"
  )) |>
  mutate(exp_gender=case_when(
    exp_gender== 0 ~ "Male", 
    exp_gender== 1 ~ "Female", 
    exp_gender== 2 ~ "Non-binary third gender", 
    exp_gender== 3 ~ "Prefer not to say"
  )) |>
  mutate(exp_race=case_when(
    exp_race== 1 ~ "American Indian / Alaskan Native", 
    exp_race== 2 ~ "Asian", 
    exp_race== 3 ~ "Black / African American", 
    exp_race== 4 ~ "Native Hawaiian / Other Pacific Islander", 
    exp_race== 5 ~ "White"
  )) |>
  mutate(exp_ethnicity=case_when(
    exp_ethnicity== 1 ~ "Hispanic / Latino", 
    exp_ethnicity== 2 ~ "Non-Hispanic/Non-Latino"
  )) |>
  mutate(exp_age_bin=case_when(
    exp_age_bin== 0 ~ "<20", 
    exp_age_bin== 20 ~ "21-25", 
    exp_age_bin== 25  ~ "26-30", 
    exp_age_bin== 30 ~ ">30"
  ))


dim(covid)
```

```{r}
coviddata <- covid2 |>
  mutate(exp_already_vax=as.factor(case_when(
    exp_already_vax== 0 ~ "No", exp_already_vax== 1 ~ "Yes"
  ))) |>
  mutate(exp_flu_vax=as.factor(case_when(
    exp_flu_vax== 0 ~ "No", exp_flu_vax== 1 ~ "Yes"
  ))) |>
  mutate(exp_profession=as.factor(case_when(
    exp_profession== 0 ~ "Medical", exp_profession== 1 ~ "Nursing"
  ))) |>
  mutate(exp_gender=as.factor(case_when(
    exp_gender== 0 ~ "Male", 
    exp_gender== 1 ~ "Female", 
    exp_gender== 2 ~ "Non-binary third gender", 
    exp_gender== 3 ~ "Prefer not to say"
  ))) |>
  mutate(exp_race=as.factor(case_when(
    exp_race== 1 ~ "American Indian / Alaskan Native", 
    exp_race== 2 ~ "Asian", 
    exp_race== 3 ~ "Black / African American", 
    exp_race== 4 ~ "Native Hawaiian / Other Pacific Islander", 
    exp_race== 5 ~ "White"
  ))) |>
  mutate(exp_ethnicity=as.factor(case_when(
    exp_ethnicity== 1 ~ "Hispanic / Latino", 
    exp_ethnicity== 2 ~ "Non-Hispanic/Non-Latino"
  ))) |>
  mutate(exp_age_bin=as.factor(case_when(
    exp_age_bin== 0 ~ "<20", 
    exp_age_bin== 20 ~ "21-25", 
    exp_age_bin== 25  ~ "26-30", 
    exp_age_bin== 30 ~ ">30"
  )))

covid_survey_longer <- coviddata |>
  pivot_longer(
    cols = starts_with("exp_"),
    names_to = "explanatory",
    values_to = "explanatory_value"
  ) |>
  filter(!is.na(explanatory_value)) |>
  pivot_longer(
    cols = starts_with("resp_"),
    names_to = "response",
    values_to = "response_value"
  )
```

```{r}
tibble(covid_survey_longer)
```

```{r}
covid_survey_summary_stats_by_group <- covid_survey_longer |> 
  mutate(response_value = as.numeric(response_value)) |>
  group_by(explanatory, explanatory_value, response) |> 
  summarise(
    mean=mean(response_value, na.rm=TRUE),
    low=quantile(response_value, .1, na.rm=TRUE),
    high=quantile(response_value,.9, na.rm=TRUE)  )

covid_survey_summary_stats_by_group
```

-   

-   Now group the data (`covid_survey_longer`) again, this time only by `response`, in order to calculate the same summary statistics for each response variable (mean, 10th percentile, and 90th percentile), not conditioned on the explanatory variables. Name this new data frame `covid_survey_summary_stats_all`. It should look like the following:

```{r}
covid_survey_summary_stats_all <- covid_survey_longer |> 
  group_by(response) |> 
  summarise(
    mean=mean(response_value, na.rm=TRUE),
    low=quantile(response_value, .1, na.rm=TRUE),
    high=quantile(response_value,.9, na.rm=TRUE)  ) |> 
  mutate(
    explanatory = "All",
    explanatory_value = ""
  )


covid_survey_summary_stats_all
```

-   Bind the two data frames of summary statistics you created `covid_survey_summary_stats_all` and `covid_survey_summary_stats_by_group` together by row. Name the resulting data frame `covid_survey_summary_stats`. It should look like the following:

```{r}
covid_survey_summary_stats <- bind_rows(covid_survey_summary_stats_all, covid_survey_summary_stats_by_group)

covid_survey_summary_stats
```

```         
```

## 4 - COVID survey - re-reconstruct

```{r fig.width=10, fig.height=5}




csss <- covid_survey_summary_stats
csss <- csss |> mutate(
    explanatory = factor(explanatory),
    explanatory_value = factor(explanatory_value),
    response = factor(response)
  ) |>
  droplevels()


p <- ggplot(csss, aes(y=explanatory_value)) +
  geom_segment(aes(x=low, xend=high), linewidth=.3) + 
  theme(
      legend.position=NULL,
      axis.ticks.x=element_blank(),
      axis.ticks.y=element_blank(), 
      panel.background = element_blank()
    ) + 
    geom_point(aes(x=low), size=.5) + 
    geom_point(aes(x=high), size=.5) + 
    geom_point(aes(x=mean), size=.5) + 
  labs(x="", y="")

xlabels <- c(
  "resp_safety" = "Based on my understanding, I believe the vaccine is safe",
"resp_confidence_science" = "I am confident in the scientific vetting process for the new COVID vaccines", 
"resp_feel_safe_at_work" = "Getting the vaccine will make me feel safer at work",
"resp_will_recommend" = "I will recommend the vaccine to family, friends, and community members",
"resp_trust_info" = "I trust the information that I have received about the vaccines",
"resp_concern_safety" = "I am concerned about the safety and side effects of the vaccine"
)

ylabels <- c(
  "exp_age_bin" = "Age",
"exp_gender" = "Gender",
"exp_race" = "Race",
"exp_ethnicity" = "Ethnicity",
"exp_profession" = "Profession",
"exp_already_vax" = "Had COVID vaccine",
"exp_flu_vax" = "Had flu vaccine this year"
)

p + facet_grid(vars(explanatory), vars(response), scales="free_y", space="free_y", drop=TRUE,
               labeller = labeller(explanatory = ylabels, response=xlabels)) + 
  theme(
    strip.text.x = element_text(size=6),
    strip.text.y = element_text(angle=0),
    strip.background = element_rect(color = "black", size = .2)
  )
```

## 5 - COVID survey - another view

**COVID survey - another view.** Create two bar charts of the Likert data for the six survey questions in from the plot in Exercise 2. This should be a single plot visualizing the percentages of each possible answer, with different questions on the y-axis. Use an appropriate color scale.

a.  Create a diverging bar chart. Write alt text for your visualization as well.

```{r}
likertdata <- covid_survey_longer |> select(response, response_value)
likertdata <- likertdata |> mutate(
  response = factor(response),
  response_value = factor(response_value)
)

countofeach <- likertdata  |> group_by(response) |> count(response_value)

# likert(likertdata)
```

```{r}
percentlikerts <- likertdata |> group_by(response, response_value) |>
  summarise(count=n()) |>
  mutate( prop = count / sum(count) )



pivoted <- pivot_wider(percentlikerts |> select(response, response_value, prop), names_from=response_value, values_from=prop)
```

```{r}
alt_text <- "This graph is a diverging horizontal bar chart of likert scores for a covid related survey. On the x-axis is the count of each likert, and the y-axis contains the different survey questions. There are 6 questions responded to: the likelihood to recommend a vaccine, to trust information regarding the vaccine, tovaccine safety, ability to feel safe at work, confidence in science, and concern for safety. Most value responses tend toward lower values, with most values in the 1-2 range. Responses on the concern for safety present a high trend, with most values either a 4 or 5. "
csl <- covid_survey_longer
ggplot(csl, aes(fill=response_value, y=response, group=response_value)) +
  geom_bar(position="diverging") +
  theme(
    panel.background = element_blank(),
    
  ) + labs(
    title="Likert scores of Covid related survey", 
    y="", x="Relative counts per survey response",
    alt=alt_text
    
  ) + 
  guides(fill=guide_legend(title="Response Values"))
```

```{r}

ggplot(percentlikerts, aes(x=response, y=prop, fill=response_value)) + 
  geom_col() +
  geom_text(aes(label = paste0(round(prop * 100, 0), "%")),
            position = position_stack(vjust = 0.5),  # place in the middle 
            color = "black",
            fontface = "bold") + 
    coord_flip() +
  labs(
    title="Likert scores of Covid related survey",
    x="", y="Percentages of survey responses"
  ) +   guides(fill=guide_legend(title="Response Values")) + 
scale_fill_brewer(palette = "Blues") + 
  theme(
    panel.background=element_blank()
  )
```
