---
title: "HW 03"
author: "Lucas Smith"
format:
  html:
    embed-resources: true
toc: true
---

```{r}
pkgs = c("tidyverse", "grid", "gridExtra", "ggplot2", "reshape")
for (pkg in pkgs) {
  if (!require(pkg, character.only=TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```

## 1 - Du Bois challenge.

```{r}
#| label: label-me-1

dubois <- read_csv("data/income.csv")
db <- pivot_longer(dubois, 
                   c("Rent", "Food", "Clothes", "Tax", "Other"), 
                   names_to="income_type", values_to = "percent")
order <- rev(c("$100-200", "$200-300", "$300-400", "$400-500", "$500-750", "$750-1000", "$1000 AND OVER"))
x_ordering <- c("Rent", "Food", "Clothes", "Tax", "Other")
# might work ggplot(db, aes(x=factor(percent, levels=x_ordering), y=Class, fill=income_type)) +  geom_col(position="fill") + 
color_vals = c("Rent" = "black", "Food" = "purple", "Clothes" = "orange", "Tax"="deepskyblue", "Other"="#97FFFF")

plt <- ggplot(db, aes(x=percent, y=Class, fill=factor(income_type, levels=rev(x_ordering)))) +  geom_col(position="fill")  
plt <- plt + 
        guides(fill=FALSE) + 
        theme(axis.text.x=element_blank(),
             axis.title.x=element_blank(),
             axis.ticks.x=element_blank(), 
             axis.ticks.y = element_blank(),
             legend.position = NULL, 
             plot.margin=margin(0,0,0,0),
             plot.background = element_blank(),
             axis.text.y = element_blank(),
             panel.background=element_blank()) + 
        labs(x="", y="") + 
        scale_y_discrete(limits=order) +
        scale_fill_manual(values=color_vals) 


tabledata <- dubois |> select(Class, Average_Income)
tab = tableGrob(tabledata, rows=NULL, theme=ttheme_default(base_size = 7, ))
tab$heights <- unit(rep(1/nrow(tab), nrow(tab) ), "npc")

grid.arrange(
  arrangeGrob(tab, nullGrob(), heights=c(200, 5)),
  arrangeGrob(nullGrob(), plt, 
              heights=c(4, 30)),
  widths=c(2, 8)
)
```

## 2 - COVID survey - interpret

```{r}
#| label: label-me-2

```

## 3 - COVID survey - reconstruct

```{r}
covid <- read_csv("data/covid-survey.csv")
```

```{r}
covid <- read_csv("data/covid-survey.csv", skip=1)

```

```{r}
# filter out any no longer useful
colnames <- colnames(covid[, -1])
filtered_missing_rows <- covid |> filter(!if_all(.cols=colnames, is.na))
dim(filtered_missing_rows)
```

-   Relabel the survey response values according to the information in `data/covid_survey.pdf`. Peeking at the plot you’re working towards reconstructing can also be helpful to identify how exactly to recode the variables. Print the dimensions of the resulting data frame with `dim()`.

    -   `exp_already_vax` and `exp_flu_vax`: 0 - No, 1 - Yes

    -   `exp_profession`: 0 - Medical, 1 - Nursing

    -   `exp_gender`: 0 - Male, 1 - Female, 3 - Non-binary third gender, 4 - Prefer not to say

    -   `exp_race`: 1 - American Indian / Alaskan Native, 2 - Asian, 3 - Black / African American, 4 - Native Hawaiian / Other Pacific Islander, 5 - White

    -   `exp_ethnicity`: 1 - Hispanic / Latino, 2 - Non-Hispanic/Non-Latino

    -   `exp_age_bin`: 0 - \<20, 20 - 21-25, 25 - 26-30, 30 - \>30

```{r}
survey_response_values <- c(
  "1" = "Strongly agree",
  "2" = "Somewhat agree",
  "3" = "Neither agree nor disagree",
  "4" = "Somewhat disagree",
  "5" = "Strongly disagree"
)

relabeled_cols <- colnames(covid[, seq(from=9, to=14)])

covid <- covid |> mutate(across(relabeled_cols, ~ survey_response_values[as.character(.)]))

covid <- covid |>
  mutate(exp_already_vax=case_when(
    exp_already_vax== 0 ~ "No", exp_already_vax== 1 ~ "Yes"
  )) |>
  mutate(exp_flu_vax=case_when(
    exp_flu_vax== 0 ~ "No", exp_flu_vax== 1 ~ "Yes"
  )) |>
  mutate(exp_profession=case_when(
    exp_profession== 0 ~ "Medical", exp_profession== 1 ~ "Nursing"
  )) |>
  mutate(exp_gender=case_when(
    exp_gender== 0 ~ "Male", 
    exp_gender== 1 ~ "Female", 
    exp_gender== 2 ~ "Non-binary third gender", 
    exp_gender== 3 ~ "Prefer not to say"
  )) |>
  mutate(exp_race=case_when(
    exp_race== 1 ~ "American Indian / Alaskan Native", 
    exp_race== 2 ~ "Asian", 
    exp_race== 3 ~ "Black / African American", 
    exp_race== 4 ~ "Native Hawaiian / Other Pacific Islander", 
    exp_race== 5 ~ "White"
  )) |>
  mutate(exp_ethnicity=case_when(
    exp_ethnicity== 1 ~ "Hispanic / Latino", 
    exp_ethnicity== 2 ~ "Non-Hispanic/Non-Latino"
  )) |>
  mutate(exp_age_bin=case_when(
    exp_age_bin== 1 ~ "<20", 
    exp_age_bin== 20 ~ "21-25", 
    exp_age_bin== 25  ~ "26-30", 
    exp_age_bin== 30 ~ ">30"
  ))


dim(covid)
```

## 4 - COVID survey - re-reconstruct

## 5 - COVID survey - another view
